---
title: "TRANSLATOR Hits for generated opposites"
author: "KBC"
date: "3/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The data

Data: I took terms from PATO and added oppositional prefixes.  Some occur natively in PATO already, some don't.  Then I did a quoted search of PubMed Central for all of them---non-negated, "native" negated, and generated.

Question: what happens if we try to validate generated negatives by seeing whether or not they actually get used?

```{r cars}
hits <- read.table("/Users/kevincohen/Dropbox/N-Z/translator-concept-oppositeness/experimental-outputs/TRANSLATOR generated antonyms - Sheet1 (1).tsv", header = TRUE, sep = "\t")

hits$native <- as.factor(hits$native.or.not)
hits$prefix <- as.factor(hits$prefix)
library(ggplot2)
ggplot(data = hits, mapping = aes(x = native, y = hits)) +
  geom_col()
ggplot(data = hits, mapping = aes(x = prefix, y = hits)) +
  geom_col()
ggplot(data = subset(hits$prefix != "∅", mappin1g = aes(x = prefix, y = hits)))

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
