---
title: "TRANSLATOR Hits for generated opposites"
author: "KBC"
date: "3/15/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## The data

Data: I took terms from PATO and added oppositional prefixes.  Some occur natively in PATO already, some don't.  Then I did a quoted search of PubMed Central for all of them---non-negated, "native" negated, and generated.

Question: what happens if we try to validate generated negatives by seeing whether or not they actually get used?

```{r cars, eval = FALSE, echo = FALSE}
hits <- read.table("/Users/kevincohen/Dropbox/N-Z/translator-concept-oppositeness/experimental-outputs/TRANSLATOR generated antonyms - Sheet1 (1).tsv", header = TRUE, sep = "\t")

hits$native <- as.factor(hits$native.or.not)
hits$prefix <- as.factor(hits$prefix)
library(ggplot2)
ggplot(data = hits, mapping = aes(x = native, y = hits)) +
  geom_col()
ggplot(data = hits, mapping = aes(x = prefix, y = hits)) +
  geom_col()
ggplot(data = subset(hits$prefix != "∅", mappin1g = aes(x = prefix, y = hits)))

```

## Comparison with Chris's paper as of 2021-03-31

1. Wrote a script that converts Chris's id-id-logical-textual format to id-id. Required more mucking about than one might have expected, but: whatever.
2. For Chris's textual-evidence-only, logical-evidence-only, and both-kinds-of-evidence pairs, I used the UNIX command _comm_ (thanks, Bill!) to find the size of the union of what he found with what I found.

The y-axis in these graphs is scaled to the total number of pairs found by all three of Chris's methods combined: 438.

The number for _K_ (what we are getting as of today) is constant across all three graphs at 105 pairs.

```{r}
we.found.textual <- 105
chris.found.textual <- 200
overlap.textual <- 0

found.by <- as_tibble(rbind(
c("TEXT", "K", 105),
c("TEXT", "C", 200),
c("TEXT", "UNION", 0)
))
# housekeeping: column names and types
colnames(found.by) <- c("evidence.from", "method", "count")
found.by$evidence.from <- as.factor(found.by$evidence.from)
found.by$method <- as.factor(found.by$method)
found.by$count <- as.integer(found.by$count)

ggplot(data = found.by, mapping = aes(x = method, y = count)) +
  geom_bar(stat = "identity") +
  ylim(0, 438) +
  labs(title = "Found by textual evidence only")

found.by <- as_tibble(rbind(
c("TEXT", "K", 105),
c("LOGICAL", "C", 52),
c("N/A", "UNION", 0)
))
# housekeeping: column names and types
colnames(found.by) <- c("evidence.from", "method", "count")
found.by$evidence.from <- as.factor(found.by$evidence.from)
found.by$method <- as.factor(found.by$method)
found.by$count <- as.integer(found.by$count)

ggplot(data = found.by, mapping = aes(x = method, y = count)) +
  geom_bar(stat = "identity") +
  ylim(0, 438) +
  labs(title = "Found by (Chris's) logical evidence only")

found.by <- as_tibble(rbind(
c("TEXT", "K", 105),
c("BOTH", "C", 184),
c("N/A", "UNION", 0)
))
# housekeeping: column names and types
colnames(found.by) <- c("evidence.from", "method", "count")
found.by$evidence.from <- as.factor(found.by$evidence.from)
found.by$method <- as.factor(found.by$method)
found.by$count <- as.integer(found.by$count)

ggplot(data = found.by, mapping = aes(x = method, y = count)) +
  geom_bar(stat = "identity") +
  ylim(0, 438) +
  labs(title = "Found by both types of Chris's evidence")

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
